---
title: "Propensity Score Estimation with ML"
format: 
  html:
    embed-resources: true
    message: false
    warning: false
    highlight-style: pygments
editor_options: 
  chunk_output_type: console
---

Instead of tuning models the usual way, we tune directly on the covariate balance achieved when weighting on the inverse of the predicted class probabilities of a classifier.

For each model, we end up with one configuration that yielded the lowest remaining covariate imbalance.

# Packages & Data

```{r}
library(cobalt)
library(discrim)
library(furrr)
library(future)
library(progressr)
library(purrr)
library(tidymodels)

# for plots:
library(showtext)
library(sysfonts)

font_add(
  "Gyre",
  regular = "/usr/share/texmf/fonts/opentype/public/tex-gyre/texgyretermes-regular.otf"
)

showtext_auto()

theme_paper <- function(base_size = 14, base_family = "Gyre") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      panel.grid.major = element_line(color = "#E5E5E5", linewidth = 0.4),
      panel.grid.minor = element_blank(),

      axis.line = element_line(color = "black"),
      axis.ticks = element_line(color = "black"),

      axis.text = element_text(color = "black"),
      axis.title = element_text(color = "black"),

      plot.title = element_text(face = "bold", size = rel(1.2)),
      plot.subtitle = element_text(color = "black"),

      legend.position = "right",
      legend.title = element_text(face = "bold"),

      panel.background = element_blank(),
      plot.background = element_blank()
    )
}

theme_set(theme_paper())

set.seed(42)

fac_to_int <- function(x) {
  as.integer(as.character(x))
}
```

# Prep

Not using settlement features for now (measured at current location = post-treatment; need to match wartime locations which might get messy). We are using respondent characteristics:

* Age
* Sex
* Ethnicity
* Experienced other victimization

```{r}
survey <- 
  readr::read_csv(here::here("data/survey.csv")) |> 
  drop_na(crsv_self_family, resp_ethn) |>
  fastDummies::dummy_cols(c("resp_ethn", "resp_educ")) |> 
  select(
    uuid, starts_with(c("resp_ethn", "resp_educ_")), resp_age, 
    crsv_self_family, victim_other, resp_ethn,
    resp_male
  ) |> 
  # outcome has to be factor for classification task with tidymodels,
  # first level is interpreted as event level ("treated" or here "moderated")
  mutate(crsv_self_family = factor(crsv_self_family, levels = c(1, 0)))
```

# Inspect

Feature correlations:

```{r}
survey |> 
  select(-c(uuid, crsv_self_family, resp_ethn)) |> 
  cor() |> 
  corrplot::corrplot(method = "color", type = "upper", tl.cex = 0.6)
```

Correlations with treatment:

```{r}
survey |> 
  select(crsv_self_family, where(is.numeric)) |> 
  summarise(
    across(
      -crsv_self_family,
      \(x) cor(x, fac_to_int(crsv_self_family), use = "complete.obs")
    )
  ) |> 
  pivot_longer(everything(), names_to = "variable", values_to = "r") |> 
  arrange(desc(abs(r)))
```

# Models & Workflow

* Random forest with tuned hyperparameters
* Penalized logistic regression with benchmark; tune both mixture (type) and penalty (amount) of regularization.

Data is imbalanced, but we won't use oversampling or SMOTE to not alter the distributions (fit model on observed data generating process). Preprocessing recipe:

```{r}
prop_rec <- 
  recipe(crsv_self_family ~ ., data = survey) |> 
  update_role(uuid, new_role = "ID") |> 
  step_normalize(all_numeric_predictors()) |> 
  # drop resp_ethn (categorical) for modelling, but keep inside
  # data for later. Also, drop one of the dummies each bc of 
  # collinearity (serve as baseline):
  step_rm(resp_ethn, resp_ethn_Croat, resp_educ_None)

# logistic regression also gets first-order interactions between all
# respondent attributes as features:
prop_rec_log <- 
  prop_rec |> 
  step_interact(terms = ~ starts_with("resp"):starts_with("resp"))
```

Models specifications. Since the logistic regression will include all first-order interaction terms, many of which may be irrelevant, we allow regularization (tuning type and strength):

```{r}
log_spec <- 
  logistic_reg(
    penalty = tune(),
    mixture = tune() # 0 = ridge, 1 = lasso, 0 < mixture < 1 = elastic net
  ) |> 
  set_engine("glmnet") |> 
  set_mode("classification")

xgb_spec <-
  boost_tree(
    trees = tune(),
    tree_depth = tune(),
    learn_rate = tune(),
    min_n = tune(),
    loss_reduction = tune(),
    sample_size = tune(),
    mtry = tune()
  ) |>
  set_engine(
    "xgboost", 
    objective = "binary:logistic", 
    counts = FALSE
  ) |>
  set_mode("classification")

rf_spec <-
  rand_forest(
    trees = tune(),
    mtry = tune(),
    min_n = tune()
  ) |>
  set_engine(
    "ranger",
    probability = TRUE,
    importance = "impurity"
  ) |>
  set_mode("classification")
```

Hyperparameter grids for tuning. The ranges are not apriori, but are derived from prior testing.

```{r}
log_grid <-
  grid_space_filling(
    penalty(range = c(-3, -1)),  # log10 scale
    mixture(range = c(0, 1)),
    size = 50
  )

xgb_grid <-
  grid_space_filling(
    trees(range = c(500, 1500)),
    tree_depth(range = c(1, 5)),
    learn_rate(range = c(1e-3, .1), trans = NULL),
    min_n(range = c(3, 30)),
    loss_reduction(range = c(1e-3, .99), trans = NULL),
    sample_size = sample_prop(range = c(.7, .9)),
    mtry = mtry_prop(range = c(.6, 1)),
    size = 200,
    type = "latin_hypercube"
  )

rf_grid <-
  grid_space_filling(
    trees(range = c(500, 2000)),
    mtry = mtry_prop(range = c(0.3, 1)),
    min_n(range = c(5, 150)),
    size = 150,
    type = "latin_hypercube"
  )
```

# Helper functions for tuning

Functions to compute balance statistics from fitted tidymodels workflow:

```{r}
augment_ipw <- function(wflow, data, clip = 0) {
  form <- formula(extract_recipe(wflow))
  lhs <- rlang::f_lhs(form)
  
  df_fit <- 
    wflow |> 
    augment(data) |> 
    rename(prop = .pred_1) |> 
    mutate(!!lhs := fac_to_int(!!lhs))
  
  true_prop <- mean(pull(df_fit, lhs), na.rm = TRUE)
  
  df_fit |> 
    mutate(
      prop = pmin(pmax(prop, clip), 1 - clip),
      ipw = if_else(
        !!lhs == 1, 
        true_prop / prop, 
        (1 - true_prop) / (1 - prop)
      )
    )
}

balance_stats <- function(formula, data, weights) {
  balance <- bal.tab(
    formula, 
    weights = weights,
    data = data, 
    estimand = "ATE"
  )
  
  tibble(
    mean_asmd = mean(abs(balance$Balance$Diff.Adj)),
    max_asmd = max(abs(balance$Balance$Diff.Adj)),
    ess_treat = balance$Observations["Adjusted", ]$Treated,
    ess_cont = balance$Observations["Adjusted", ]$Control
  )
}
```

# Tuning

For every model, fit every config from the grid to the data, calculate stabilized inverse probability weights & compute remaining imbalance. We are trying to find the model/config that generates the best balance.

Model configs:

```{r}
models <- tribble(
  ~id,             ~spec,     ~recipe,      ~grid,
  "Random Forest", rf_spec,   prop_rec,     rf_grid,
  "XGBoost",       xgb_spec,  prop_rec,     xgb_grid,
  "Elastic Net",   log_spec,  prop_rec_log, log_grid,
)

models
```

Tuning procedure. Parallelizing over all but one available cores to speed this up.

```{r}
# Using constant formula since `prop_rec_log` includes interaction terms
# that tidymodels builds into the df (means the formula extracted from
# that recipe will have columns that do not exist in the original data):
FORMULA <- formula(prep(prop_rec, survey))

plan(multisession, workers = parallel::detectCores() - 1)

tuning_res <- pmap(models, function(id, spec, recipe, grid) {
  
  n <- nrow(grid)
  
  message(sprintf("\nFitting %d models for %s...", n, id))
  
  with_progress({
    
    p <- progressor(steps = n)
    
    # iterate over grid:
    
    future_map(seq(n), function(i) {
      
      p()
      
      # fit each model config. to the data...
      
      hyperparams <- grid[i, ]
      
      wf <- 
        workflow() |> 
        add_recipe(recipe) |> 
        add_model(update(spec, hyperparams)) |> 
        fit(survey)
        
      df_fit <- wf |> augment_ipw(survey)
      
      # ...and compute balance:
      bal <- balance_stats(FORMULA, df_fit, df_fit$ipw)
      bind_cols(tibble(id = id, config = list(hyperparams)), bal)
        
    }, .options = furrr_options(seed = 42)) |> bind_rows()
    
  })
  
}) |> bind_rows()
```

## Tuning results

Also fitting a plain logistic regression on the prepped data with all interaction terms, as benchmark (in practice, researchers often resort to using logistic regression to estimate propensity scores, so this is a reasonable benchmark):

```{r}
plain_log <- 
  logistic_reg() |> 
  set_engine("glm") |> 
  set_mode("classification")
  
plain_log_wf <- 
  workflow() |> 
  add_recipe(prop_rec_log) |> # `prop_rec_log` adds first-order interactions
  add_model(plain_log)

plain_log_fit <- 
  plain_log_wf |> 
  fit(survey) |> 
  augment_ipw(survey) |> 
  (\(df) balance_stats(FORMULA, df, df$ipw))() |> 
  bind_cols(id = "Logistic Regression")
```

Best balance achieved by each model, by lowest mean ASMD:

```{r}
tuning_res |> 
  group_by(id) |> 
  arrange(mean_asmd, .by_group = TRUE) |> 
  slice(1) |> 
  ungroup() |> 
  bind_rows(plain_log_fit) |> 
  arrange(mean_asmd)
```

Distribution of mean ASMDs by model:

```{r}
tuning_res |> 
  ggplot(aes(x = mean_asmd, y = id, color = id, fill = id)) +
  geom_vline(xintercept = .1, lty = "dotted") +
  geom_violin(alpha = .25) +
  labs(fill = "Model", color = "Model", y = "Model", x = "Mean ASMD") +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2")

ggsave(here::here("figs/mean_asmd_by_model.pdf"), width = 7, height = 5)
```

Max ASMDs:

```{r}
tuning_res |> 
  ggplot(aes(x = max_asmd, y = id, color = id, fill = id)) +
  geom_vline(xintercept = .1, lty = "dotted") +
  geom_violin(alpha = .25) +
  labs(fill = "Model", color = "Model", y = "Model", x = "Max. ASMD") +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2")

ggsave(here::here("figs/max_asmd_by_model.pdf"), width = 7, height = 5)
```

ESS vs. mean ASMD:

```{r}
tuning_res |> 
  ggplot(aes(x = ess_treat, y = mean_asmd, color = id)) +
  geom_hline(yintercept = .1, lty = "dotted") +
  geom_point(alpha = .34) +
  geom_smooth(se = FALSE, linewidth = .5) +
  labs(x = "ESS Treated Units", y = "Mean ASMD", color = "Model") +
  scale_color_brewer(palette = "Dark2")

ggsave(here::here("figs/ess_vs_mean_asmd.pdf"), width = 7, height = 5)
```

ESS vs. max ASMD

```{r}
tuning_res |> 
  ggplot(aes(x = ess_treat, y = max_asmd, color = id)) +
  geom_hline(yintercept = .1, lty = "dotted") +
  geom_point(alpha = .34) +
  geom_smooth(se = FALSE, linewidth = .5) +
  labs(x = "ESS Treated Units", y = "Max. ASMD", color = "Model") +
  scale_color_brewer(palette = "Dark2")

ggsave(here::here("figs/ess_vs_max_asmd.pdf"), width = 7, height = 5)
```

Mean vs. Max ASMD (checking for "sacrifice one"-behavior):

```{r}
tuning_res |> 
  ggplot(aes(x = max_asmd, y = mean_asmd, color = id)) +
  geom_vline(xintercept = .1, lty = "dotted") +
  geom_hline(yintercept = .1, lty = "dotted") +
  geom_point(alpha = .34) +
  geom_smooth(se = FALSE, linewidth = .5) +
  labs(x = "Max. ASMD", y = "Mean ASMD", color = "Model") +
  scale_color_brewer(palette = "Dark2")

ggsave(here::here("figs/mean_vs_max_asmd.pdf"), width = 7, height = 5)
```

# Hyperparams

Exporting hyperparameter plots: for each model, how does each hyperparameter impact mean ASMD?

```{r}
models |> 
  pull(id) |> 
  map(function(model) {
    
    name <- tolower(gsub(" ", "_", model))
    
    suppressWarnings({
    
      p <- 
        tuning_res |> 
        filter(id == model) |> 
        unnest(config) |> 
        select(id:mean_asmd) |> 
        pivot_longer(
          cols = -c(id, mean_asmd), 
          names_to = "param", 
          values_to = "value"
        ) |> 
        ggplot(aes(x = value, y = mean_asmd)) +
        geom_point(alpha = .5) +
        geom_smooth() +
        facet_wrap(~param, scales = "free_x") +
        labs(x = "Value", y = "Mean ASMD")
      
    })
    
    ggsave(
      here::here(glue::glue("figs/hyperparams_{name}.pdf")),
      plot = p,
      width = 7,
      height = 5
    )
    
    p
})

# ^ could use `walk()` to save but I wanted to show the plots in the HTML
```

# Export

Getting the best config for each model:

```{r}
models_full <- 
  tuning_res |> 
  group_by(id) |> 
  arrange(mean_asmd, .by_group = TRUE) |> 
  slice(1) |> 
  ungroup() |> 
  select(id, config) |> 
  right_join(models)

models_full |> pull(config) |> set_names(models_full$id)
```

Fitting the best models to the data & getting estimates, so that for each unit we have an estimated propensity score & IPW from each best model:

```{r}
estimates <- 
  models_full |> 
  mutate(
    
    wflow = map2(recipe, spec, function(rec, mod) {
      workflow() |> 
        add_recipe(rec) |> 
        add_model(mod)
    }),
    
    final = map2(wflow, config, \(wf, cfg) finalize_workflow(wf, cfg)),
    
    fit = map(final, function(wf) {
      wf |> 
        fit(survey) |> 
        augment_ipw(survey)
    })
    
  ) |> 
  select(id, fit) |> 
  unnest(fit) |> 
  select(id, uuid, ipw, prop, everything())

estimates
```

Exporting:

```{r}
estimates |> 
  select(id, uuid, ipw, prop) |> 
  readr::write_csv(here::here("data/propensity_scores.csv"))
```

# Additional diagnostics

Overlap:

```{r}
estimates |> 
  mutate(crsv_self_family = factor(crsv_self_family, levels = c(1, 0))) |> 
  ggplot(aes(x = prop, color = crsv_self_family, fill = crsv_self_family)) +
  geom_density(alpha = .25) +
  facet_wrap(~id, ncol = 2) +
  theme(legend.position = "bottom") +
  labs(
    x = "P(S=1|X)",
    y = "Density",
    fill = "S",
    color = "S"
  ) +
  scale_color_brewer(palette = "Dark2", direction = -1) +
  scale_fill_brewer(palette = "Dark2", direction = -1)

ggsave(here::here("figs/overlap_by_model.pdf"), width = 7, height = 5)
```

Balance:

```{r}
balances <- 
  estimates |> 
  group_by(id) |> 
  nest() |> 
  ungroup() |> 
  mutate(
    balance = map(data, function(df) {
      
      tab <- bal.tab(
        FORMULA,
        weights = df$ipw,
        data = df,
        estimand = "ATE",
        un = TRUE
      )$Balance
      
      tab |> 
        tibble::rownames_to_column("covariate") |> 
        rename_with(\(n) tolower(gsub("\\.", "_", n)))
      
    })
  ) |> 
  select(id, balance) |> 
  unnest(balance)
```

Visually:

```{r}
balances |> 
  select(-type) |> 
  pivot_longer(starts_with("diff"), names_to = "adj", values_to = "diff") |> 
  mutate(
    adj = adj |> replace_values(
      "diff_un" ~ "Unadjusted",
      "diff_adj" ~ "Adjusted"
    ),
    adj = factor(adj, levels = c("Unadjusted", "Adjusted")),
    diff = abs(diff)
  ) |> 
  ggplot(
    aes(
      x = diff, 
      y = forcats::fct_reorder(covariate, diff), 
      color = adj
    )
  ) +
  geom_vline(xintercept = 0) +
  geom_vline(xintercept = .1, lty = "dotted") +
  geom_point() +
  facet_wrap(~id, ncol = 2) +
  labs(x = "SMD", y = "", color = "") +
  theme(legend.position = "bottom") +
  scale_color_brewer(palette = "Dark2", direction = -1) +
  scale_fill_brewer(palette = "Dark2", direction = -1)

ggsave(here::here("figs/final_balance_love_plot.pdf"), width = 7, height = 5)
```

Effective Sample Size:

```{r}
estimates |> 
  group_by(id) |> 
  nest() |> 
  ungroup() |> 
  mutate(
    balance = map(data, function(df) {
      
      obs <- bal.tab(
        FORMULA,
        weights = df$ipw,
        data = df,
        estimand = "ATE",
        un = TRUE
      )$Observations
      
      obs |> tibble::rownames_to_column("adj")
      
    })
  ) |> 
  select(id, balance) |> 
  unnest(balance) |> 
  mutate(adj = factor(adj, levels = c("Unadjusted", "Adjusted"))) |> 
  pivot_longer(c(Control, Treated), names_to = "treat", values_to = "val") |> 
  ggplot(aes(x = treat, y = val, color = adj, fill = adj)) +
  geom_col(position = position_dodge(width = .95), alpha = .5) +
  facet_wrap(~id, ncol = 2) +
  theme(legend.position = "bottom") +
  labs(x = "", y = "Effective Sample Size", color = "", fill = "") +
  scale_color_brewer(palette = "Dark2", direction = -1) +
  scale_fill_brewer(palette = "Dark2", direction = -1)

ggsave(here::here("figs/final_balance_ess.pdf"), width = 7, height = 5)
```
